Is partner choice related to prosociality? A cross-national investigation

### 2022-09-12 manuscript.tex
added new supp figures and tables

\PassOptionsToPackage{unicode=true}{hyperref} % options for packages loaded elsewhere
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames*,x11names*}{xcolor}
%
\documentclass[english,man,floatsintext]{apa6}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provides euro and other symbols
\else % if luatex or xelatex
  \usepackage{unicode-math}
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage[]{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\usepackage{xcolor}
\usepackage{hyperref}
\hypersetup{
            pdftitle={Is partner choice related to prosociality? A cross-national investigation},
            pdfauthor={Scott Claessens1 \& Thanos Kyritsis1},
            pdfkeywords={partner choice, relational mobility, cooperation, prosociality, cross-cultural},
            colorlinks=true,
            linkcolor=Maroon,
            filecolor=Maroon,
            citecolor=Blue,
            urlcolor=blue,
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}

% set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

% Manuscript styling
\usepackage{upgreek}
\captionsetup{font=singlespacing,justification=justified}

@@

% Create new environments so endfloat can handle them
% \newenvironment{ltable}
%   {\begin{landscape}\begin{center}\begin{threeparttable}}
%   {\end{threeparttable}\end{center}\end{landscape}}
\newenvironment{lltable}{\begin{landscape}\begin{center}\begin{ThreePartTable}}{\end{ThreePartTable}\end{center}\end{landscape}}

@@

\linenumbers
\usepackage{csquotes}
\raggedbottom
\usepackage{setspace}
\AtBeginEnvironment{tabular}{\singlespacing}
\AtBeginEnvironment{lltable}{\singlespacing}
\AtBeginEnvironment{tablenotes}{\doublespacing}
\captionsetup[table]{font={stretch=1.5}}
\captionsetup[figure]{font={stretch=1.5}}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[shorthands=off,main=english]{babel}
\else
  % load polyglossia as late as possible as it *could* call bidi if RTL lang (e.g. Hebrew or Arabic)
  \usepackage{polyglossia}
  \setmainlanguage[]{english}
\fi

@@

One factor that could explain global variation in prosociality is differing possibilities for partner choice across countries. Here, \enquote{partners} are defined as individuals that people socially interact with to provide mutual benefits (e.g.~friends, neighbours, colleagues, mates). Theoretical models of partner choice show that when individuals can leave interactions with uncooperative partners and actively choose new interactions with cooperative partners, cooperation can evolve and be sustained (Aktipis, 2004, 2011; Enquist \& Leimar, 1993; Roberts, 2020, 1998; Roberts et al., 2021). Partner choice allows for the assortative matching of cooperators, creating a market in which individuals use prosocial displays to compete for access to profitable social partnerships (Barclay, 2013, 2016). Thus, partner choice models predict that humans should be more prosocial and cooperative if they are able to leave unprofitable partnerships and freely choose new partnerships.

@@

As well as predicting behaviour in the lab and in small-scale societies, partner choice models also predict that socioecological conditions favouring partner choice should promote prosociality in countries around the world. One recently developed socioecological variable that captures differing possibilities for partner choice is relational mobility (Yuki \& Schug, 2012). Relational mobility captures \enquote{how much freedom and opportunity a society affords individuals to choose and dispose of interpersonal relationships based on personal preference} (p.~7521; Thomson et al., 2018). In societies with low relational mobility, people do not actively choose their relationships and their social partners are relatively fixed. By contrast, in societies with high relational mobility, people actively choose who they interact with, creating dynamic social networks. The former is akin to classic partner control models in evolutionary game theory, where individuals are forced to interact for a fixed period (Axelrod \& Hamilton, 1981), whereas the latter is akin to models of partner choice and biological markets (Barclay, 2013).

@@

Altruism was measured by two items; a hypothetical charitable donation (\enquote{Imagine the following situation: Today you unexpectedly received 1000 euros. How much of this amount would you donate to a good cause?}) and willingness to unconditionally donate to charity (\enquote{How willing are you to give to good causes without expecting anything in return?}). Trust was measured by a single item: agreement with the statement \enquote{I assume that people have only the best intentions}. Positive reciprocity was measured by two items: stating the price of a hypothetical thank-you gift the participant would give to a stranger who helped them, and agreement with the statement \enquote{When someone does me a favour I am willing to return it}. These items have been shown to reliably predict altruistic, trusting, and reciprocal behaviour in incentivised economic decision-making experiments (Falk et al., 2016).

@@

In addition to our main variables, we also included several control variables in our regressions. These control variables are justified by a causal model in which both relational mobility and prosociality are jointly affected by various confounds (see Supplementary Figure \ref{fig:plotDAG}).

@@

Second, we controlled for geographic and linguistic proximity between countries. Countries that are close to one another and share common cultural ancestors are likely to be more similar to one another, due to similar ecologies, climates, institutions, and norms (see Supplementary Figure \ref{fig:plotDAG}). To account for these unmeasured confounds, we allowed countries to covary according to geographic and linguistic proximity in our models. Geographic proximity was calculated as the inverse of the logged geodesic distance between country capital cities (data from the R package \emph{maps}; Brownrigg, 2018) using the R package \emph{geosphere} (Hijmans, 2019). Linguistic proximity between two countries was calculated as the cultural proximity between all languages spoken within those countries, weighted by speaker percentages (see Supplementary Methods for more details; Hammarström et al., 2017; Eberhard et al., 2018).

@@

Participants in both the World Values Survey and the European Values Survey answer a range of self-report questions on social values, societal wellbeing, trust, economic values, religion, politics, and ethics. For the purposes of our study, we highlighted several variables as measures of cooperation, trust, and prosociality. The first variable captures cooperation via collective action (``Are you a member of a charitable or humanitarian organisation?''; for a similar interpretation of this variable, see Jacquet et al., 2021). The second variable captures generalised trust (\enquote{Generally speaking, would you say that most people can be trusted or that you need to be very careful in dealing with people?}). The third set of variables captures levels of trust in specific groups of people, namely family, neighborhood, personal acquaintances, people the respondent has met for the first time, people of another religion, and people of another nationality. The fourth set of variables captures the justifiability of different self-interested moral trangressions, including claiming unentitled government benefits, avoiding a fare on public transport, cheating on taxes, and someone accepting a bribe.

@@

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/plotWVS6a-1.pdf}
\caption{\label{fig:plotWVS6a}\emph{Posterior predictions from a Bayesian multilevel logistic regression predicting generalised trust from country-level relational mobility, without controls.} The line and shaded area indicate the median posterior regression line and 95\% credible intervals. Points indicate the proportion of individuals stating that \enquote{most people can be trusted} on the y-axis and relational mobility scores on the x-axis, for each of the 32 countries, with error bars representing +/- 1 standard error.}
\end{figure}

@@

Overall, contrary to our pre-registered hypotheses, we found that relational mobility was unrelated to collective action (operationalised as charitable organisation membership), generalised trust, and moral justifiability ratings for self-interested behaviours. Relational mobility was also unrelated to trust in most specific groups, though we did find that relational mobility negatively predicted trust in family and positively predicted trust in people of another religion and nationality. This \enquote{scope of trust} effect, whereby relational mobility is associated with lower trust in closer contacts but greater trust in more distant contacts, is an interesting feature of the construct that aligns with previous work (Thomson et al., 2018).

@@

Why did we not find the expected relationships between relational mobility and prosociality for most measures? One might argue that relational mobility is not an adequate measure of the kinds of partner choice implemented in theoretical models of cooperation or laboratory experiments. We would contest this view. Relational mobility is explicitly defined as a construct that quantifies \enquote{variance in partner choice in human societies} (p.~7521; Thomson et al., 2018). In the relational mobility scale, people are asked about their immediate society, including friends and acquaintances, and whether people in this immediate society can \enquote{leave {[}current relationships{]} for better ones} and \enquote{choose\ldots{} the people they interact with}. These are the exact same opportunities afforded to agents in partner choice models and participants in partner choice experiments.

@@

As well as raising questions about previous work on the evolution of cooperation in humans, our findings also add to a growing body of evidence linking relational mobility to trust. We found a \enquote{scope of trust} effect, whereby relational mobility negatively predicts trust in close contacts (family members) and positively predicts trust in distant contacts (people of other religions and nationalities). This finding builds on previous work finding that relational mobility predicts trust in strangers (Thomson et al., 2018) and additionally shows, with multiple groups of increasing social distance, that relational mobility scales up people's circles of trust beyond close kin. However, we also found that relational mobility was not reliably related to generalised trust across our two studies. This result contradicts previous reported associations between relational mobility and generalised trust (Yuki et al., 2007; Yuki \& Schug, 2012).

@@

\hypertarget{refs}{}
\leavevmode\hypertarget{ref-Aktipis2004}{}%
Aktipis, C. A. (2004). Know when to walk away: Contingent movement and the evolution of cooperation. \emph{Journal of Theoretical Biology}, \emph{231}(2), 249--260. \url{https://doi.org/https://doi.org/10.1016/j.jtbi.2004.06.020}

\leavevmode\hypertarget{ref-Aktipis2011}{}%
Aktipis, C. A. (2011). Is cooperation viable in mobile organisms? Simple walk away rule favors the evolution of cooperation in groups. \emph{Evolution and Human Behavior}, \emph{32}(4), 263--276. \url{https://doi.org/10.1016/j.evolhumbehav.2011.01.002}

\leavevmode\hypertarget{ref-Aust2020}{}%
Aust, F., \& Barth, M. (2020). \emph{papaja: Prepare reproducible APA journal articles with R Markdown}. \url{https://github.com/crsh/papaja}

\leavevmode\hypertarget{ref-Axelrod1981}{}%
Axelrod, R., \& Hamilton, W. D. (1981). The evolution of cooperation. \emph{Science}, \emph{211}(4489), 1390--1396. \url{https://doi.org/10.1126/science.7466396}

\leavevmode\hypertarget{ref-Barclay2004}{}%
Barclay, P. (2004). Trustworthiness and competitive altruism can also solve the "tragedy of the commons". \emph{Evolution and Human Behavior}, \emph{25}(4), 209--220. \url{https://doi.org/10.1016/j.evolhumbehav.2004.04.002}

\leavevmode\hypertarget{ref-Barclay2013}{}%
Barclay, P. (2013). Strategies for cooperation in biological markets, especially for humans. \emph{Evolution and Human Behavior}, \emph{34}(3), 164--175. \url{https://doi.org/10.1016/j.evolhumbehav.2013.02.002}

\leavevmode\hypertarget{ref-Barclay2016b}{}%
Barclay, P. (2016). Biological markets and the effects of partner choice on cooperation and friendship. \emph{Current Opinion in Psychology}, \emph{7}, 33--38. \url{https://doi.org/10.1016/j.copsyc.2015.07.012}

\leavevmode\hypertarget{ref-Barclay2016a}{}%
Barclay, P., \& Raihani, N. (2016). Partner choice versus punishment in human prisoner's dilemmas. \emph{Evolution and Human Behavior}, \emph{37}(4), 263--271. \url{https://doi.org/10.1016/j.evolhumbehav.2015.12.004}

\leavevmode\hypertarget{ref-Barclay2007}{}%
Barclay, P., \& Willer, R. (2007). Partner choice creates competitive altruism in humans. \emph{Proceedings of the Royal Society B: Biological Sciences}, \emph{274}(1610), 749--753. \url{https://doi.org/10.1098/rspb.2006.0209}

\leavevmode\hypertarget{ref-BliegeBird2015}{}%
Bliege Bird, R., \& Power, E. A. (2015). Prosocial signaling and cooperation among Martu hunters. \emph{Evolution and Human Behavior}, \emph{36}(5), 389--397. \url{https://doi.org/10.1016/j.evolhumbehav.2015.02.003}

\leavevmode\hypertarget{ref-Bromham2018}{}%
Bromham, L., Hua, X., Cardillo, M., Schneemann, H., \& Greenhill, S. J. (2018). Parasites and politics: Why cross-cultural studies must control for relatedness, proximity and covariation. \emph{Royal Society Open Science}, \emph{5}(8), 181100.

\leavevmode\hypertarget{ref-Brownrigg2018}{}%
Brownrigg, R. (2018). \emph{maps: Draw geographical maps}. \url{https://CRAN.R-project.org/package=maps}

\leavevmode\hypertarget{ref-Burkner2017}{}%
Bürkner, P.-C. (2017). brms: An R package for Bayesian multilevel models using Stan. \emph{Journal of Statistical Software}, \emph{80}(1), 1--28. \url{https://doi.org/10.18637/jss.v080.i01}

\leavevmode\hypertarget{ref-Cronk2019}{}%
Cronk, L., Berbesque, C., Conte, T., Gervais, M., Iyer, P., McCarthy, B., Sonkoi, D., Townsend, C., \& Aktipis, A. (2019). Managing risk through cooperation: Need-based transfers and risk pooling among the societies of the Human Generosity Project. In L. R. Lozny \& T. H. McGovern (Eds.), \emph{Global perspectives on long term community resource management} (pp. 41--75). Springer International Publishing. \url{https://doi.org/10.1007/978-3-030-15800-2_4}

\leavevmode\hypertarget{ref-Dorrough2016}{}%
Dorrough, A. R., \& Glöckner, A. (2016). Multinational investigation of cross-societal cooperation. \emph{Proceedings of the National Academy of Sciences}, \emph{113}(39), 10836--10841. \url{https://doi.org/10.1073/pnas.1601294113}

\leavevmode\hypertarget{ref-Ethnologue}{}%
Eberhard, D. M., Simons, G. F., \& Fennig, C. D. (Eds.). (2018). \emph{Ethnologue: Languages of the world} (Twenty-first). SIL International.

\leavevmode\hypertarget{ref-Enquist1993}{}%
Enquist, M., \& Leimar, O. (1993). The evolution of cooperation in mobile organisms. \emph{Animal Behaviour}, \emph{45}(4), 747--757. \url{https://doi.org/10.1006/anbe.1993.1089}

\leavevmode\hypertarget{ref-Falk2018}{}%
Falk, A., Becker, A., Dohmen, T., Enke, B., Huffman, D., \& Sunde, U. (2018). Global evidence on economic preferences. \emph{The Quarterly Journal of Economics}, \emph{133}(4), 1645--1692. \url{https://doi.org/10.1093/qje/qjy013}

\leavevmode\hypertarget{ref-Falk2016}{}%
Falk, A., Becker, A., Dohmen, T., Huffman, D. B., \& Sunde, U. (2016). The preference survey module: A validated instrument for measuring risk, time, and social preferences. \emph{IZA Discussion Papers}, \emph{9674}.

\leavevmode\hypertarget{ref-Glottolog}{}%
Hammarström, H., Forkel, R., Haspelmath, M., \& Bank, S. (2017). \emph{Glottolog 3.0}. Max Planck Institute for the Science of Human History. \url{https://doi.org/10.5281/zenodo.4061162}

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  man,floatsintext]{apa6}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\ifLuaTeX
\usepackage[bidi=basic]{babel}
\else
\usepackage[bidi=default]{babel}
\fi
\babelprovide[main,import]{english}
% get rid of language-specific shorthands (see #6817):
\let\LanguageShortHands\languageshorthands
\def\languageshorthands#1{}
% Manuscript styling
\usepackage{upgreek}
\captionsetup{font=singlespacing,justification=justified}

@@

% Create new environments so endfloat can handle them
% \newenvironment{ltable}
%   {\begin{landscape}\centering\begin{threeparttable}}
%   {\end{threeparttable}\end{landscape}}
\newenvironment{lltable}{\begin{landscape}\centering\begin{ThreePartTable}}{\end{ThreePartTable}\end{landscape}}

@@

\linenumbers
\usepackage{csquotes}
\raggedbottom
\usepackage{setspace}
\AtBeginEnvironment{tabular}{\singlespacing}
\AtBeginEnvironment{lltable}{\singlespacing}
\AtBeginEnvironment{tablenotes}{\doublespacing}
\captionsetup[table]{font={stretch=1.5}}
\captionsetup[figure]{font={stretch=1.5}}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Is partner choice related to prosociality? A cross-national investigation},
  pdfauthor={Scott Claessens1 \& Thanos Kyritsis1},
  pdflang={en-EN},
  pdfkeywords={partner choice, relational mobility, cooperation, prosociality, cross-cultural},
  colorlinks=true,
  linkcolor={Maroon},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={blue},
  pdfcreator={LaTeX via pandoc}}

@@

One factor that could explain global variation in prosociality is differing possibilities for partner choice across countries. Here, `partners' are defined as individuals that people socially interact with to provide mutual benefits (e.g.~friends, neighbours, colleagues, mates). Theoretical models of partner choice show that when individuals can leave interactions with uncooperative partners and actively choose new interactions with cooperative partners, cooperation can evolve and be sustained (Aktipis, 2004, 2011; Enquist \& Leimar, 1993; Roberts, 2020, 1998; Roberts et al., 2021). Partner choice allows for the assortative matching of cooperators, creating a market in which individuals use prosocial displays to compete for access to profitable social partnerships (Barclay, 2013, 2016). Thus, partner choice models predict that humans should be more prosocial and cooperative if they are able to leave unprofitable partnerships and freely choose new partnerships.

@@

As well as predicting behaviour in the lab and in small-scale societies, partner choice models also predict that socioecological conditions favouring partner choice should promote prosociality in countries around the world. One recently developed socioecological variable that captures differing possibilities for partner choice is relational mobility (Yuki \& Schug, 2012). Relational mobility captures ``how much freedom and opportunity a society affords individuals to choose and dispose of interpersonal relationships based on personal preference'' {[}p.~7521; Thomson et al. (2018){]}. In societies with low relational mobility, people do not actively choose their relationships and their social partners are relatively fixed. By contrast, in societies with high relational mobility, people actively choose who they interact with, creating dynamic social networks. The former is akin to classic partner control models in evolutionary game theory, where individuals are forced to interact for a fixed period (Axelrod \& Hamilton, 1981), whereas the latter is akin to models of partner choice and biological markets (Barclay, 2013).

@@

Altruism was measured by two items; a hypothetical charitable donation (``Imagine the following situation: Today you unexpectedly received 1000 euros. How much of this amount would you donate to a good cause?'') and willingness to unconditionally donate to charity (``How willing are you to give to good causes without expecting anything in return?''). Trust was measured by a single item: agreement with the statement ``I assume that people have only the best intentions''. Positive reciprocity was measured by two items: stating the price of a hypothetical thank-you gift the participant would give to a stranger who helped them, and agreement with the statement ``When someone does me a favour I am willing to return it''. These items have been shown to reliably predict altruistic, trusting, and reciprocal behaviour in incentivised economic decision-making experiments (Falk et al., 2016).

@@

In addition to our main variables, we also included several control variables in our regressions. These control variables are justified by a causal model in which both relational mobility and prosociality are jointly affected by various confounds (see Figure \ref{fig:plotDAG}).

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/plotDAG-1.pdf}
\caption{\label{fig:plotDAG}\emph{Directed acyclic graph of the causal model justifying the inclusion of covariates in our statistical models.} Thomson et al.~(2018) show that environmental harshness (Harsh) and subsistence style (Sub) are antecedents of relational mobility (RM), but other evidence also suggests that environmental harshness and subsistence style directly affect prosociality (Pro; Cronk et al., 2019; Talhelm et al., 2014). Environmental harshness and subsistence style are thus third variables that confound the direct path from relational mobility to prosociality. Moreover, all four of these variables are confounded by unmeasured factors (U), such as ecology, climate, institutions, and norms. We cannot directly condition on unmeasured factors, but since these factors are themselves predicted by geographic (Geo) and linguistic (Lin) proximity between countries, we can account for them by allowing countries to covary according to geographic and linguistic proximity.}
\end{figure}

@@

Second, we controlled for geographic and linguistic proximity between countries. Countries that are close to one another and share common cultural ancestors are likely to be more similar to one another, due to similar ecologies, climates, institutions, and norms (see Figure \ref{fig:plotDAG}). To account for these unmeasured confounds, we allowed countries to covary according to geographic and linguistic proximity in our models. Geographic proximity was calculated as the inverse of the logged geodesic distance between country capital cities {[}data from the R package \emph{maps}; Brownrigg (2018){]} using the R package \emph{geosphere} (Hijmans, 2019). Linguistic proximity between two countries was calculated as the cultural proximity between all languages spoken within those countries, weighted by speaker percentages {[}see Supplementary Methods for more details; Hammarström et al. (2017); Eberhard et al. (2018){]}.

@@

Participants in both the World Values Survey and the European Values Survey answer a range of self-report questions on social values, societal wellbeing, trust, economic values, religion, politics, and ethics. For the purposes of our study, we highlighted several variables as measures of cooperation, trust, and prosociality. The first variable captures cooperation via collective action {[}``Are you a member of a charitable or humanitarian organisation?''; for a similar interpretation of this variable, see Jacquet et al. (2021){]}. The second variable captures generalised trust (``Generally speaking, would you say that most people can be trusted or that you need to be very careful in dealing with people?''). The third set of variables captures levels of trust in specific groups of people, namely family, neighborhood, personal acquaintances, people the respondent has met for the first time, people of another religion, and people of another nationality. The fourth set of variables captures the justifiability of different self-interested moral trangressions, including claiming unentitled government benefits, avoiding a fare on public transport, cheating on taxes, and someone accepting a bribe.

@@

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/plotWVS6a-1.pdf}
\caption{\label{fig:plotWVS6a}\emph{Posterior predictions from a Bayesian multilevel logistic regression predicting generalised trust from country-level relational mobility, without controls.} The line and shaded area indicate the median posterior regression line and 95\% credible intervals. Points indicate the proportion of individuals stating that ``most people can be trusted'' on the y-axis and relational mobility scores on the x-axis, for each of the 32 countries, with error bars representing +/- 1 standard error.}
\end{figure}

@@

Overall, contrary to our pre-registered hypotheses, we found that relational mobility was unrelated to collective action (operationalised as charitable organisation membership), generalised trust, and moral justifiability ratings for self-interested behaviours. Relational mobility was also unrelated to trust in most specific groups, though we did find that relational mobility negatively predicted trust in family and positively predicted trust in people of another religion and nationality. This ``scope of trust'' effect, whereby relational mobility is associated with lower trust in closer contacts but greater trust in more distant contacts, is an interesting feature of the construct that aligns with previous work (Thomson et al., 2018).

@@

Why did we not find the expected relationships between relational mobility and prosociality for most measures? One might argue that relational mobility is not an adequate measure of the kinds of partner choice implemented in theoretical models of cooperation or laboratory experiments. We would contest this view. Relational mobility is explicitly defined as a construct that quantifies ``variance in partner choice in human societies'' {[}p.~7521; Thomson et al. (2018){]}. In the relational mobility scale, people are asked about their immediate society, including friends and acquaintances, and whether people in this immediate society can ``leave {[}current relationships{]} for better ones'' and ``choose\ldots{} the people they interact with''. These are the exact same opportunities afforded to agents in partner choice models and participants in partner choice experiments.

@@

As well as raising questions about previous work on the evolution of cooperation in humans, our findings also add to a growing body of evidence linking relational mobility to trust. We found a ``scope of trust'' effect, whereby relational mobility negatively predicts trust in close contacts (family members) and positively predicts trust in distant contacts (people of other religions and nationalities). This finding builds on previous work finding that relational mobility predicts trust in strangers (Thomson et al., 2018) and additionally shows, with multiple groups of increasing social distance, that relational mobility scales up people's circles of trust beyond close kin. However, we also found that relational mobility was not reliably related to generalised trust across our two studies. This result contradicts previous reported associations between relational mobility and generalised trust (Yuki et al., 2007; Yuki \& Schug, 2012).

@@

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-Aktipis2004}{}}%
Aktipis, C. A. (2004). Know when to walk away: Contingent movement and the evolution of cooperation. \emph{Journal of Theoretical Biology}, \emph{231}(2), 249--260. https://doi.org/\url{https://doi.org/10.1016/j.jtbi.2004.06.020}

\leavevmode\vadjust pre{\hypertarget{ref-Aktipis2011}{}}%
Aktipis, C. A. (2011). Is cooperation viable in mobile organisms? Simple walk away rule favors the evolution of cooperation in groups. \emph{Evolution and Human Behavior}, \emph{32}(4), 263--276. \url{https://doi.org/10.1016/j.evolhumbehav.2011.01.002}

\leavevmode\vadjust pre{\hypertarget{ref-Aust2020}{}}%
Aust, F., \& Barth, M. (2020). \emph{{papaja}: {Prepare} reproducible {APA} journal articles with {R Markdown}}. \url{https://github.com/crsh/papaja}

\leavevmode\vadjust pre{\hypertarget{ref-Axelrod1981}{}}%
Axelrod, R., \& Hamilton, W. D. (1981). The evolution of cooperation. \emph{Science}, \emph{211}(4489), 1390--1396. \url{https://doi.org/10.1126/science.7466396}

\leavevmode\vadjust pre{\hypertarget{ref-Barclay2004}{}}%
Barclay, P. (2004). Trustworthiness and competitive altruism can also solve the "tragedy of the commons". \emph{Evolution and Human Behavior}, \emph{25}(4), 209--220. \url{https://doi.org/10.1016/j.evolhumbehav.2004.04.002}

\leavevmode\vadjust pre{\hypertarget{ref-Barclay2013}{}}%
Barclay, P. (2013). Strategies for cooperation in biological markets, especially for humans. \emph{Evolution and Human Behavior}, \emph{34}(3), 164--175. \url{https://doi.org/10.1016/j.evolhumbehav.2013.02.002}

\leavevmode\vadjust pre{\hypertarget{ref-Barclay2016b}{}}%
Barclay, P. (2016). Biological markets and the effects of partner choice on cooperation and friendship. \emph{Current Opinion in Psychology}, \emph{7}, 33--38. \url{https://doi.org/10.1016/j.copsyc.2015.07.012}

\leavevmode\vadjust pre{\hypertarget{ref-Barclay2016a}{}}%
Barclay, P., \& Raihani, N. (2016). Partner choice versus punishment in human prisoner's dilemmas. \emph{Evolution and Human Behavior}, \emph{37}(4), 263--271. \url{https://doi.org/10.1016/j.evolhumbehav.2015.12.004}

\leavevmode\vadjust pre{\hypertarget{ref-Barclay2007}{}}%
Barclay, P., \& Willer, R. (2007). Partner choice creates competitive altruism in humans. \emph{Proceedings of the Royal Society B: Biological Sciences}, \emph{274}(1610), 749--753. \url{https://doi.org/10.1098/rspb.2006.0209}

\leavevmode\vadjust pre{\hypertarget{ref-BliegeBird2015}{}}%
Bliege Bird, R., \& Power, E. A. (2015). Prosocial signaling and cooperation among {Martu} hunters. \emph{Evolution and Human Behavior}, \emph{36}(5), 389--397. \url{https://doi.org/10.1016/j.evolhumbehav.2015.02.003}

\leavevmode\vadjust pre{\hypertarget{ref-Bromham2018}{}}%
Bromham, L., Hua, X., Cardillo, M., Schneemann, H., \& Greenhill, S. J. (2018). Parasites and politics: Why cross-cultural studies must control for relatedness, proximity and covariation. \emph{Royal Society Open Science}, \emph{5}(8), 181100.

\leavevmode\vadjust pre{\hypertarget{ref-Brownrigg2018}{}}%
Brownrigg, R. (2018). \emph{{m}aps: Draw geographical maps}. \url{https://CRAN.R-project.org/package=maps}

\leavevmode\vadjust pre{\hypertarget{ref-Burkner2017}{}}%
Bürkner, P.-C. (2017). {brms}: An {R} package for {Bayesian} multilevel models using {Stan}. \emph{Journal of Statistical Software}, \emph{80}(1), 1--28. \url{https://doi.org/10.18637/jss.v080.i01}

\leavevmode\vadjust pre{\hypertarget{ref-Cronk2019}{}}%
Cronk, L., Berbesque, C., Conte, T., Gervais, M., Iyer, P., McCarthy, B., Sonkoi, D., Townsend, C., \& Aktipis, A. (2019). Managing risk through cooperation: Need-based transfers and risk pooling among the societies of the {Human Generosity Project}. In L. R. Lozny \& T. H. McGovern (Eds.), \emph{Global perspectives on long term community resource management} (pp. 41--75). Springer International Publishing. \url{https://doi.org/10.1007/978-3-030-15800-2_4}

\leavevmode\vadjust pre{\hypertarget{ref-Dorrough2016}{}}%
Dorrough, A. R., \& Glöckner, A. (2016). Multinational investigation of cross-societal cooperation. \emph{Proceedings of the National Academy of Sciences}, \emph{113}(39), 10836--10841. \url{https://doi.org/10.1073/pnas.1601294113}

\leavevmode\vadjust pre{\hypertarget{ref-Ethnologue}{}}%
Eberhard, D. M., Simons, G. F., \& Fennig, C. D. (Eds.). (2018). \emph{Ethnologue: Languages of the world} (Twenty-first). SIL International.

\leavevmode\vadjust pre{\hypertarget{ref-Enquist1993}{}}%
Enquist, M., \& Leimar, O. (1993). The evolution of cooperation in mobile organisms. \emph{Animal Behaviour}, \emph{45}(4), 747--757. \url{https://doi.org/10.1006/anbe.1993.1089}

\leavevmode\vadjust pre{\hypertarget{ref-Falk2018}{}}%
Falk, A., Becker, A., Dohmen, T., Enke, B., Huffman, D., \& Sunde, U. (2018). Global evidence on economic preferences. \emph{The Quarterly Journal of Economics}, \emph{133}(4), 1645--1692. \url{https://doi.org/10.1093/qje/qjy013}

\leavevmode\vadjust pre{\hypertarget{ref-Falk2016}{}}%
Falk, A., Becker, A., Dohmen, T., Huffman, D. B., \& Sunde, U. (2016). The preference survey module: A validated instrument for measuring risk, time, and social preferences. \emph{IZA Discussion Papers}.
